<!-- # --- -->
title: "preprocessing cohen data"
output: html_notebook
---

obsidian://open?vault=science&file=Excalidraw%2Fcohen_data_map_LKGLKarSVFqsg0mh0TID7KgPed2HxVRM

libs & init definitions
```{r}
data_root<-"/home/mcn26/palmer_scratch/tabula_data"
output_root="/home/mcn26/palmer_scratch/tabula_data/formatted"
#whether data should be automatically downloaded : run notebook once with true,
#then subsequently with false
download_data=FALSE
library(tidyverse)
library(Seurat)
```
Let's get the cohen data. Specifically, let's start with the mouse retina dataset. 
(my apologies for the R. Will be useful if we decide to play with the endogenous
transcriptome & need to load seurat objects). 

# Download data

https://zenodo.org/records/7338678/files/scMPRA.zip?download=1

```{r}
zenodo=paste(data_root,"/scMPRA.zip",sep="")
unzipped_zenodo_path=paste(data_root,"/scMPRA",sep="")
if(download_data){
  download.file("https://zenodo.org/records/7338678/files/scMPRA.zip?download=1",zenodo)
  dir.create(unzipped_zenodo_path)
  unzip(zenodo,exdir=unzipped_zenodo_path)
  file.remove(zenodo)
}
```

# Load retina data

## Load and process MPRAbc data

```{r}
retina_MPRAbc_rep1<-read_csv(paste(data_root,"/scMPRA/scMPRA/data/sc_expression_retina/retina_pBC_ec_rBC_ec_rep1.csv",sep=""))
retina_MPRAbc_rep2<-read_csv(paste(data_root,"/scMPRA/scMPRA/data/sc_expression_retina/retina_pBC_ec_rBC_ec_rep2.csv",sep=""))
```
Preferable to have the two replicates in one tibble
```{r}
retina_MPRAbc_rep1 <- retina_MPRAbc_rep1 %>% mutate(rep_id=1)
retina_MPRAbc_rep2 <- retina_MPRAbc_rep2 %>% mutate(rep_id=2)
retina_MPRAbc <- bind_rows(retina_MPRAbc_rep1, retina_MPRAbc_rep2)
```

Kill the original variables to free some space. 
```{r}
rm(retina_MPRAbc_rep1,retina_MPRAbc_rep2)
gc()
```



```{r}
retina_MPRAbc
```

Interestingly there are some zeroes un the umi name and count. Are zeroes in 
these columns always coincident?

```{r}
table(
  Count = ifelse(retina_MPRAbc$count == 0, "count == 0", "count != 0"),
  UMI = ifelse(retina_MPRAbc$umi == "0", "umi == '0'", "umi != '0'")
)
```

Perfect overlap : no UMI means no count and vice versa. Any NAs to mess things up?

```{r}
any(is.na(retina_MPRAbc$count) | is.na(retina_MPRAbc$umi))
```

Ok, none are NA, perfect.

We want UMI counts, not counts of reads of individual UMIs, so let's aggregate. 
```{r}
retina_MPRAbc_agg <- retina_MPRAbc %>%
  group_by(cellBC, rBC, name, pBC, cluster, rep_id) %>%
  summarise(
    total_count = sum(count),
    UMIs_MPRA_BC = n_distinct(umi[count > 0]),
    .groups = "drop"
  )
```


## Loading transfection reporter (U6) data

Let's factor in the U6 data. 

There are two versions of the U6 data. One is labeled "padded". Examining the 
tables, the non-padded version appears to be discrete, whole-number read counts, so we'll use that. 


```{r}
u6_retina_r1<-read_csv(paste(data_root,"/scMPRA/scMPRA/data/sc_expression_retina/retina_u6_pBC_umi_rep1.csv",sep=""))
u6_retina_r2<-read_csv(paste(data_root,"/scMPRA/scMPRA/data/sc_expression_retina/retina_u6_pBC_umi_rep2.csv",sep=""))
```

Preferable to have the two replicates in one tibble
```{r}
u6_retina_r1 <- u6_retina_r1 %>% mutate(rep_id=1)
u6_retina_r2 <- u6_retina_r2 %>% mutate(rep_id=2)
u6_retina <- bind_rows(u6_retina_r1, u6_retina_r2)
```

Kill the original variables to free some space. 
```{r}
rm(u6_retina_r1,u6_retina_r2)
gc()
```

Are any zero?

```{r}
any(u6_retina$counts==0)
```
none, good. 

Since there is no pBC<->promoter name association in that table, let's check to 
see if all of the reported pBCs map to promoters...
First, we will extract the pBC<->promoter name map:

```{r}
map=unique(retina_MPRAbc_agg[c("name","pBC")])
```

Make sure that the map is 1:1 :
```{r}
stopifnot(nrow(unique(map["name"]))==nrow(unique(map["pBC"])))
stopifnot(nrow(unique(map["name"]))==nrow(map))
```

are there any pBCs in the u6 data which are not real?

```{r}
all(u6_retina[["pBC"]] %in% map[["pBC"]])
```

No, they are all real (in the map). 


## Adding transfection reporter (U6) data

Let's add the U6 counts to the MPRAbc tibble
First, let's unify the names

```{r}
u6_retina=u6_retina %>%
  rename(
    cluster=clustrer,
    u6_count=counts
  )
```


Next, we'll join the two dataframes.
I will perform an outer join to fill in zeroes: transfected but not expressed, 
and expressed but did not detect transfection reporter. 

Note that one effect of this step will be the duplication of many transfection 
reporter counts. Since one transfection reporter has a one-to-many relationship 
with many MPRA barcodes, there's not a great way to avoid this.
(consequence of cohen construct design & limitations of csvs)
(ideally id keep in 2x tables and link, but this is csvs not sql). 
Since we don't really care about the specific U6 count numbers (just presence or 
absence) I think this is OK. 

First, though, let's make sure that the "key" is unique in the U6 table. 
If it's not, it could cause unpleasant duplication of MPRAbc data. 

```{r}
key=c('cellBC','pBC','rep_id')
```

```{r}
stopifnot(nrow(unique(u6_retina[key]))==nrow(u6_retina[key]))
```
Ok good. 

outer join
```{r}
joined=merge(retina_MPRAbc_agg, u6_retina, by=c('cellBC','pBC','rep_id'), all=TRUE)
```

Let's take a look at how the join went. It's an outer join, so we shouldn't lose
any information, but there is the possibility of introducing NAs. 
Were there any NAs before?

```{r}
any(is.na(retina_MPRAbc_agg)) || any(is.na(u6_retina))
```
Nope!

Are there any NAs after?

```{r}
any(is.na(joined))
```

Yep. these are the aforementioned "transfected but not expressed, 
and expressed but did not detect transfection reporter". Let's fill them in with
zeroes (a more approproiate marker for "not observed" than NA, which suggests 
"not measured"). 

```{r}
joined
```


Fill in the cluster info : sometimes coming from each...
Make sure the two cols never contradict each other, and there are never any case
s where we have NO cluster information. 
```{r}
error_df <- joined %>%
  filter(is.na(cluster.x) & is.na(cluster.y) | (cluster.x != cluster.y & !is.na(cluster.x) & !is.na(cluster.y)))

stopifnot(nrow(error_df)==0)
```

make a new cluster column
```{r}
cluster_col <- ifelse(!is.na(joined$cluster.x), joined$cluster.x, joined$cluster.y)
```
make sure it's complete
```{r}
stopifnot(!any(is.na(cluster_col)))
```
add new and kill old cluster columns
```{r}
joined$cluster=cluster_col
joined=joined %>% select (-c(cluster.x, cluster.y))
```


fill some NAs
```{r}
joined=joined %>%
  mutate(u6_count = if_else(is.na(u6_count), 0, u6_count)) %>%
  mutate(rBC = if_else(is.na(rBC),"0",rBC))  %>%
  mutate(total_count = if_else(is.na(total_count),0,total_count))  %>%
  mutate(UMIs_MPRA_BC = if_else(is.na(UMIs_MPRA_BC),0,UMIs_MPRA_BC))
```

Fill in the promoter name based on the map

```{r}
rows_joined_before=nrow(joined)

joined=joined %>% 
  select (-c(name)) %>%
  left_join(map,by="pBC")

stopifnot(rows_joined_before==nrow(joined))

```



```{r}
any(is.na(joined))
```

All NAs filled!


Make sure we haven't created any strange situations where umi is zero but its 
counts aren't, or vice-versa



```{r}
invalid_rows <- joined %>%
  filter(xor(rBC=="0",UMIs_MPRA_BC==0))

#invalid_rows
stopifnot(nrow(invalid_rows)==0)
```

Perfect. 

```{r}
joined
```

Rename

```{r}
joined=joined %>% 
  rename(#new=old
    MPRA_BC=rBC,
    reads_MPRA_BC=total_count,
    cell_type_annotation=cluster,
    transfection_BC=pBC,
    CRE_id=name,
    reads_transfection_BC=u6_count
  )
```

```{r}
joined
```


re-order the columns

```{r}
joined=joined %>%
  select(cellBC,rep_id,CRE_id,cell_type_annotation,MPRA_BC,reads_MPRA_BC,UMIs_MPRA_BC,transfection_BC,reads_transfection_BC)
```


make a little 2x2 grid

```{r}
table(
  mpra_count = ifelse(joined$UMIs_MPRA_BC == 0, "undetected", "detected"),
  u6_count = ifelse(joined$reads_transfection_BC == 0, "no tfection reporter", "tfection reporter")
)
```

- detected+no tfection reporter = transfected and expressed, but we lost the reporter in RNA processing.
- detected+tfection reporter = transfected and expressed
and so forth

The only weird one is double zero: no U6 and no tfection reporter. 
No evidence for those : not even real zero counts/ let us dump them.

```{r}
joined=joined %>%
  filter(UMIs_MPRA_BC != 0 | reads_transfection_BC != 0)
```




dump to disc
```{r}
write.table(joined, file = paste(output_root,"COHEN_RETINA.tsv",sep="/"), sep = "\t", row.names = FALSE, quote = FALSE)
```



